ISYE 6501 | Submission HW2

### 2022-09-07 hw2/anshit/Homework_2.Rmd
Analysis of K-Means

---
title: "Homework_2"
output: html_document
date: "2022-09-04"
---

```{r}
# imports
library(ggplot2)
library(cowplot)
library(reshape2)
library(factoextra)
```

## Question 3.1

```{r}
cc_data = read.table('./data 3.1/credit_card_data-headers.txt', header = TRUE, sep='', dec='.')
head(cc_data)
dim(cc_data)
```

@@

```{r}
iris_data = read.table('./iris.txt', header = TRUE, sep='', dec='.')
head(iris_data,50)
dim(iris_data)
```

@@

```{r}
# Density Plots
my_plots = lapply(colnames(iris_data), function(x){
  p = ggplot(iris_data) + aes_string(x)
  if(x=='Species') {
    p = p + geom_bar()
  }
  else {
    p = p + geom_density()
  }
})

@@

```{r}
# Correlation Matrix
cor_mat = round(cor(iris_data[,1:4]), 2)
melted_mat = melt(cor_mat)
ggplot(melted_mat, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile() + geom_text(aes(Var1, Var2, label=value), color='white', size=3)
```

@@

```{r}
plot(elbow_graph_data, type='b', ylab='Total Sum of Square Within', main='Elbow Graph', col='blue')
```
```{r}
# Calculation of best 'K'
bw_graph_data <- data.frame(K=integer(), between_sum_of_square=double())
for(k in seq(1, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  bw_graph_data[nrow(bw_graph_data) + 1, ] = c(k, round(km.res$betweenss,2))
}
#options(scipen=999)
bw_graph_data
```

@@

```{r}
# Calculation of best 'K'
ratio_graph_data <- data.frame(K=integer(), ratio=double())
for(k in seq(2, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  ratio_graph_data[nrow(ratio_graph_data) + 1, ] = c(k, round(km.res$tot.withinss/km.res$betweenss,2))
}
#options(scipen=999)
ratio_graph_data
```
```{r}
plot(ratio_graph_data, type='b', ylab='Ratio Within/Between SS', main='Quality', col='red')
```

@@

```{r}
best_km <- kmeans(as.matrix(scale(iris_data[,1:4])), 3, nstart = 25)
out<-cbind(iris_data,tagged_cluster=best_km$cluster)
out
```
```{r}
boxplot(Sepal.Length~tagged_cluster,data=out)
```

@@

out$tagged_cluster<-as.factor(out$tagged_cluster)
ggplot(out, aes(x=tagged_cluster, y=Petal.Width,fill=tagged_cluster)) + geom_boxplot(alpha=0.3) +theme(legend.position="none")
```
Petal width and Petal Length are strong indicators for one class (setosa)

@@

```
```{r}
classifier_df$status<-as.factor(classifier_df$status)
ggplot(classifier_df, aes(x=status, y=Sepal.Length,fill=status)) + geom_boxplot(alpha=0.3) +theme(legend.position="none")

@@

@@

```{r}
fviz_cluster(best_km, data=scale(iris_data[,1:2]), geom=c("point"), ellipse.type = "euclid")
```
***Analysing For K=4***

@@

>>>>

---
title: "Homework_2"
output: html_document
date: "2022-09-04"
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
# imports
library(ggplot2)
library(cowplot)
library(reshape2)
library(factoextra)
```

@@

```{r, results='hide'}
iris_data = read.table('./iris.txt', header = TRUE, sep='', dec='.')
head(iris_data,50)
dim(iris_data)
```

@@

```{r}
# Density Plots
my_plots = lapply(colnames(iris_data), function(x){
  p = ggplot(iris_data, aes(fill = Species)) + aes_string(x)
  if(x!='Species') {
    p = p + geom_density(alpha=.34)
  }
})

@@

From the above distribution, we can make the following inferences: 1.
Petal.Length and Petal.Width provide clear separation between the setosa
vs the other classes. In clustering we expect these variables to be
significantly lower in the cluster dominated by the 'setosa' classes. 2.
We observe that while Sepal.Width and Sepal.Length have different
central tendencies, they also show high degree of overlap between the
different classes at the fringes.

```{r}
# Correlation Matrix
cor_mat = round(cor(iris_data[,1:4]), 2)
cor_mat[upper.tri(cor_mat)] <- NA
melted_mat = melt(cor_mat)
ggplot(melted_mat, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile() + geom_text(aes(Var1, Var2, label=value), color='white', size=3)
```

From the above matrix we observed the following: 1. Petal Length and
Width are highly correlated 2. Sepal Length and Petal Width are highly
correlated also 3. Sepal length and Sepal Width show negligible
dependence 4. Sepal width is negatively correlated petal length and
width

@@

```{r}
plot(elbow_graph_data, type='b', ylab='Total Sum of Square Within', main='Elbow Graph', col='blue')
```

From the above graph, we can see that increasing the value of K beyond 4
yields miminal benefit in clustering.

```{r}
# Calculation of best 'K'
bw_graph_data <- data.frame(K=integer(), between_sum_of_square=double())
for(k in seq(1, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  bw_graph_data[nrow(bw_graph_data) + 1, ] = c(k, round(km.res$betweenss,2))
}
#options(scipen=999)
bw_graph_data
```

@@

```{r}
# Calculation of best 'K'
ratio_graph_data <- data.frame(K=integer(), ratio=double())
for(k in seq(2, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  ratio_graph_data[nrow(ratio_graph_data) + 1, ] = c(k, round(km.res$tot.withinss/km.res$betweenss,2))
}
#options(scipen=999)
ratio_graph_data
```

```{r}
plot(ratio_graph_data, type='b', ylab='Ratio Within/Between SS', main='Quality', col='red')
```

@@

```{r}
best_km <- kmeans(as.matrix(scale(iris_data[,1:4])), 3, nstart = 25)
out<-cbind(iris_data,tagged_cluster=best_km$cluster)
out
```

From the confusion matrix, we can infer the following: \* Cluster 3
shows highest proportion of setosa class. In fact it encompasses 100% of
that class. \* Cluster 2 has the highest proportion of versicolor class.
\* Cluster 1 has the highest proportion of virginia class.

However in comparison to cluster 3 which is purely setosa; clusters 1
and 2 have a significant no. of data points of a different class. In
other words, there is confusion between classes in clusters 1 and 2.

@@

out$tagged_cluster<-as.factor(out$tagged_cluster)
ggplot(out, aes(x=tagged_cluster, y=Petal.Width,fill=tagged_cluster)) + geom_boxplot(alpha=0.3) +theme(legend.position="none") + xlab('Cluster')
```

-   As we expected from the density plots and confusion matrix, Petal
    width and Petal Length are significant discriminator for cluster 3,
    which is dominated by setosa class.
-   While the other two variables provide relatively lower
    discrimination between the clusters.

@@

```

```{r}
classifier_df$status<-as.factor(classifier_df$status)
ggplot(classifier_df, aes(x=status, y=Sepal.Length,fill=status)) + geom_boxplot(alpha=0.3) +theme(legend.position="none")

@@

From the above plots, we see: \* Distribution of sepal length is very
similar for improperly classified versicolor and proper virginica.
Similarly, the distribution of improperly classified virginica and
proper versicolor are very similar too. \* The same effect is apparent
in Sepal width. \* This apparent similarity can be a probable reason for
clustering to confuse between the classes.

@@

-   It can be seen from the plot above that cluster 1 and 2 are closely
    knit (i.e. intra cluster separation is less) and hence we have the
    confusion in clustering some points for these 2 clusters . Cluster 3
    has a better separability than the other 2 and hence it is a good
    demarcated cluster.

-   It is also apparent that cluster 3 has higher spread and hence can
    be split up into 2 clusters depending on the use case which also
    came up in our previous study (elbow graph to decide between k=3 and
    k=4)

***Analysing For K=4***

@@

```{r}
plot(scale(iris_data[,1:4]), col=alt_kmeans$cluster)
```

```{r}
fviz_cluster(alt_kmeans, data=scale(iris_data[,1:2]), geom=c("point"), ellipse.type = "euclid")
```

### How well the clustering fits the flower type ?

As seen from above the total accuracy of the clustering model is:

```{r}
1-(((11+14)/nrow(iris_data)))
```

### Accuracy in individual classes

If cluster 1 is virginica, cluster 2 is versicolor, cluster 3 is setosa

```{r}
acc_versicolor = 1 - (11/50)
acc_virginica = 1 - (14/50)
acc_setosa = 1- (0/50)

print (paste("Acc for versicolor",acc_versicolor))
print (paste("Acc for virginica",acc_virginica))
print (paste("Acc for setosa",acc_setosa))
```

Thus we have 100% accuracy for setosa, while for the other two flowers
we have accuracy in 70's.

### 2022-09-07 hw2/submission.Rmd
Analysis of K-Means

>>>>

## Question 4.1

### Example 1: Market Segmentation/Retail Marketing

Clustering models are frequently used for grouping customers with similar attributes

**Objective Function (**$Y$): **Predictor Variables (**$X_i$):

-   ***Household Income***:
-   ***Household Size***:
-   ***Distance From Nearest Urban Area***:
-   ***Head of Houehold Occupation***:

## Question 4.2

```{r, results='hide'}
iris_data = read.table('./iris.txt', header = TRUE, sep='', dec='.')
head(iris_data,50)
dim(iris_data)
```

### Exploratory Data Analysis

```{r}
summary(iris_data)
```

```{r}
table(iris_data$Species)
```

```{r}
# Density Plots
my_plots = lapply(colnames(iris_data), function(x){
  p = ggplot(iris_data, aes(fill = Species)) + aes_string(x)
  if(x!='Species') {
    p = p + geom_density(alpha=.34)
  }
})

plot_grid(plotlist = my_plots)
```

[**#Analysis**]{style="color:red"} From the above distribution, we can make the following inferences: 1. Petal.Length and Petal.Width provide clear separation between the setosa vs the other classes. In clustering we expect these variables to be significantly lower in the cluster dominated by the 'setosa' classes. 2. We observe that while Sepal.Width and Sepal.Length have different central tendencies, they also show high degree of overlap between the different classes at the fringes.

```{r}
# Correlation Matrix
cor_mat = round(cor(iris_data[,1:4]), 2)
cor_mat[upper.tri(cor_mat)] <- NA
melted_mat = melt(cor_mat)
ggplot(melted_mat, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile() + geom_text(aes(Var1, Var2, label=value), color='white', size=3)
```

[**#Analysis**]{style="color:red"} From the above matrix we observed the following: 1. Petal Length and Width are highly correlated 2. Sepal Length and Petal Width are highly correlated also 3. Sepal length and Sepal Width show negligible dependence 4. Sepal width is negatively correlated petal length and width

### Modelling

```{r}
# Calculation of best 'K' by elbow graph for total sum of squares within
elbow_graph_data <- data.frame(K=integer(), within_sum_of_square=double())
for(k in seq(1, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  elbow_graph_data[nrow(elbow_graph_data) + 1, ] = c(k, km.res$tot.withinss)
}
plot(elbow_graph_data, type='b', ylab='Total Sum of Square Within', main='Elbow Graph', col='blue')
```

[**#Analysis**]{style="color:red"} \* From the above graph, we can see that increasing the value of 'K' beyond 4 yields very low marginal benefit. \* Both 3 and 4 appear to be good values of 'K' and can be used depending on the use case. In this example, since we know that there are 3 classes, we expect 3 to be a more appropriate choice.

```{r}
# Other Method for determining best 'K' by sum of intra-cluster distance
bw_graph_data <- data.frame(K=integer(), between_sum_of_square=double())
for(k in seq(1, 10, by=1)){
  km.res <- kmeans(as.matrix(scale(iris_data[,1:4])), k, nstart = 25)
  bw_graph_data[nrow(bw_graph_data) + 1, ] = c(k, round(km.res$betweenss,2))
}
plot(bw_graph_data, type='b', ylab='Betweeen Sum of Square (Intra Cluster)', main='Betweeen Sum of Square (Intra Cluster) Graph', col='green')
```

[**#Analysis**]{style="color:red"} \* Similarly to elbow-graph, this plot also shows 3 and 4 to be good choices for 'K'.

```{r, include=FALSE}
best_km <- kmeans(as.matrix(scale(iris_data[,1:4])), 3, nstart = 25)
out<-cbind(iris_data,tagged_cluster=best_km$cluster)
out$tagged_cluster<-as.factor(out$tagged_cluster)

getBoxPlot <- function(df, y_val){
  ggplot(df, aes(x=tagged_cluster, y=y_val,fill=tagged_cluster)) + geom_boxplot(alpha=0.3) +theme(legend.position="none")  + xlab('Cluster')
}
```

```{r}
confusion_matrix <- table(iris_data$Species, best_km$cluster)
confusion_matrix
```

[**#Analysis**]{style="color:red"} From the confusion matrix, we can infer the following: \* Cluster 3 shows highest proportion of setosa class. In fact it encompasses 100% of that class. \* Cluster 2 has the highest proportion of versicolor class. \* Cluster 1 has the highest proportion of virginia class.

However in comparison to cluster 3 which is purely setosa; clusters 1 and 2 have a significant no. of data points of a different class. In other words, there is confusion between classes in clusters 1 and 2.

```{r}
getBoxPlot(out, 'Sepal.Length')
getBoxPlot(out, 'Sepal.Width')
getBoxPlot(out, 'Petal.Length')
getBoxPlot(out, 'Petal.Width')
```

[**#Analysis**]{style="color:red"} \* As we expected from the density plots and confusion matrix, Petal width and Petal Length are significant discriminator for cluster 3, which is dominated by setosa class. \* While the other two variables provide relatively lower discrimination between the clusters.

### Analysis of the misclassified points

```{r}
classifier_df = out[!out$Species=='setosa',]
classifier_df$status = ''
classifier_df$status[(classifier_df$tagged_cluster=="1" & classifier_df$Species=="versicolor")] = 'improper_versicolor'
classifier_df$status[(classifier_df$tagged_cluster=="2" & classifier_df$Species=="virginica")] = 'improper_virginica'
classifier_df$status[(classifier_df$tagged_cluster=="2" & classifier_df$Species=="versicolor")] = 'proper_versicolor'
classifier_df$status[(classifier_df$tagged_cluster=="1" & classifier_df$Species=="virginica")] = 'proper_virginica'
classifier_df$status<-as.factor(classifier_df$status)
```

```{r}
getBoxPlot(classifier_df, 'Sepal.Length')
getBoxPlot(classifier_df, 'Sepal.Width')
getBoxPlot(classifier_df, 'Petal.Length')
getBoxPlot(classifier_df, 'Petal.Width')
```

[**#Analysis**]{style="color:red"}
From the above plots, we see:
* Distribution of sepal length is very similar for improperly classified versicolor and proper virginica. Similarly, the distribution of improperly classified virginica and proper versicolor are very similar too.
* The same effect is apparent in Sepal width.
* This apparent similarity can be a probable reason for clustering to confuse between the classes.

```{r}
# Centroids of Clusters
kmeans_basic_table <- data.frame(best_km$size, best_km$centers)
kmeans_basic_table
```

### Plotting Clusters

```{r}
fviz_cluster(best_km, data=scale(iris_data[,1:2]), geom=c("point"), ellipse.type = "euclid")
```

[**#Analysis**]{style="color:red"}
* It can be seen from the plot above that cluster 1 and 2 are closely knit (i.e. intra cluster separation is less) and hence we have the confusion in clustering some points for these 2 clusters . Cluster 3 has a better separability than the other 2 and hence it is a good demarcated cluster.
* It is also apparent that cluster 3 has higher spread and hence can be split up into 2 clusters depending on the use case which also came up in our previous study (elbow graph to decide between k=3 and k=4)

### Just a visualization with 4 Cluster Space to highlight the above point

```{r}
alt_kmeans <- kmeans(as.matrix(scale(iris_data[,1:4])), 4, nstart = 25)
alt_kmeans
fviz_cluster(alt_kmeans, data=scale(iris_data[,1:2]), geom=c("point"), ellipse.type = "euclid")
```

### Accuracy in individual classes

If cluster 1 is virginica, cluster 2 is versicolor, cluster 3 is setosa

```{r}
acc_versicolor = 1 - (11/50)
acc_virginica = 1 - (14/50)
acc_setosa = 1- (0/50)

print (paste("Acc for versicolor",acc_versicolor))
print (paste("Acc for virginica",acc_virginica))
print (paste("Acc for setosa",acc_setosa))
```

[**#Analysis**]{style="color:red"}
Thus we have 100% accuracy for setosa, while for the other two flowers
we have accuracy in 70's.
